---
title: "COVID-19 and Religion"
author: "Sean McGoey"
date: "4/24/20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, paged.print=TRUE)
```

## Objective

Analyzing available data to determine if there are any possible links between particularly religious areas and COVID-19 cases. We know that COVID is highly transmissible from person to person, and religious services bring millions of people together across the country every week (or at least they used to -- and they will again soon as states start to reopen). So let's take a look at whether there's any relationship.

## Setup

First, we'll load our necessary packages.

```{r}

# Load tidyverse, tidycensus, janitor, scales, mapview, ggthemes
library(tidyverse)
library(tidycensus)
library(janitor)
library(scales)
library(mapview)
library(ggthemes)

```

Then we'll load in county-level COVID-19 data from The New York Times and data from the Association of Religion Data Archives' 2010 Religious Congregation Membership Study. (Ideally we'd have newer religion data than 2010, but the survey is conducted every 10 years.) We'll also grab shifted geodata from the census -- we'll use it to look at maps later.

```{r}

# Read in NYT county data, and build "key" column
nyt_covid_county <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>%
  mutate(key = paste(county, state, sep = "|", collapse=NULL))

# Read in 2010 Religion Census data
rcms_religion_data <- read.csv("data/rcms_2010.csv") %>%
  clean_names() %>%
  rename(congregations_per_10k = congregations_per_10k_people)

# Define Census API Key
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

# Load ACS variable definitions
acs_variables <- load_variables(2017, "acs5" )

# Pull in shifted geodata
county_geodata_shifted <- get_acs(geography = "county",
              variables = "B01001_001", geometry = TRUE, shift_geo = TRUE)

```

Next, because the NYT data is a cumulative set -- tracking the total number of cases/deaths every day rather than each day's new count -- we'll pull out only the highest values to find the most updated totals.

```{r}

# Group by key phrase in order to examine each county's case count peak to date
nyt_covid_county_max <- nyt_covid_county %>%
  filter(!is.na(fips)) %>%
  filter(!str_detect(county, "Unknown")) %>%
  group_by(key, county, state, fips) %>%
  summarise(case_peak = max(cases))

```

Then we'll join the table of max case counts to the religion census data and add a column for cases per 10K population.

```{r}

county_covid_religion <- inner_join(nyt_covid_county_max, rcms_religion_data, BY=key) %>%
  select(key, fips, county, state, population, adherents, congregations, congregations_per_10k, religious_adherence, case_peak) %>%
  mutate(cases_per_10k = case_peak / population * 10000) %>%
  mutate(fips = toString(fips)) %>%
  mutate(fips = if_else(str_length(fips) == 4, paste("0", fips, sep="", collapse = NULL), fips))

# NOTE: Mutate functions convert fips from an int to a string and solves for leading 0s being omitted when formatted as numbers.

```

We'll also build a table that allows us to look at variables on the map.

```{r}

geo_county_covid_religion <- county_geodata_shifted %>%
 inner_join(county_covid_religion, by=c("GEOID" = "fips"))

```

## First variable: Religious adherence

We're going to start by looking at the relationship between religious adherence and confirmed cases so far. Let's plot out some maps. First, per-capita COVID cases in each county across the country:

```{r}

geo_county_covid_religion %>%
  ggplot(aes(fill = cases_per_10k)) +
  geom_sf(lwd = 0) +
  theme_map() +
  labs(fill='Confirmed cases per 10,000 people', caption = "Source: New York Times COVID-19 data") +
  scale_fill_continuous(labels = comma) +
  theme(legend.position="bottom") +
  scale_fill_viridis_c(option = "plasma",labels = comma)
  
```

A few well-defined hot spots around the country, but not necessarily much in the way of a pattern. Next, we'll look at religious adherence:

```{r}

geo_county_covid_religion %>%
  ggplot(aes(fill = religious_adherence)) +
  geom_sf(lwd = 0) +
  theme_map() +
  labs(fill='% of religious adherents',caption = "Source: The Association of Religion Data Archives, 2010 survey") +
  scale_fill_continuous(labels = comma) +
  theme(legend.position="bottom") +
  scale_fill_viridis_c(option = "plasma",labels = comma)
  
```

That's a much more defined pattern: Lots of adherents in the South, Midwest and Plains states, with a few other pockets around the country.

But we need more than just maps. Let's see if we can find an actual relationship between religious adherence and COVID cases so far.

```{r}

ggplot(county_covid_religion) +
  geom_point(aes(religious_adherence, cases_per_10k)) +
  labs(x="Religious adherence %", y="Peak COVID cases per 10,000 people", caption = "Source: NYT COVID-19 data and the Association of Religion Data Archives", fill="key") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_smooth(aes(religious_adherence, cases_per_10k), method = "lm", se = FALSE)

```

It doesn't look like there's much of a relationship; we can check that with a correlation test.

```{r}

# Test the relationship
correlation_covid_adherents <- county_covid_religion %>%
  ungroup() %>%
  select(religious_adherence, cases_per_10k) 

cor.test(correlation_covid_adherents$religious_adherence, correlation_covid_adherents$cases_per_10k)

```

Okay, so there's really nothing there to suggest that there's a link between the most religious counties in the U.S. and the cases of COVID-19 that have been confirmed so far.

## Second variable: Concentration of religious adherents

What about another measure: not just the number of people who are adherents in a given county, but the *concentration* of those adherents? The ARDA data set includes a figure for congregations per 10,000 residents, which would help us assess whether areas with many different churches are more likely to have seen COVID outbreaks.

As we did with adherents, let's start with the maps. A quick reminder of the distribution of COVID cases:

```{r}

geo_county_covid_religion %>%
  ggplot(aes(fill = cases_per_10k)) +
  geom_sf(lwd = 0) +
  theme_map() +
  labs(fill='Confirmed cases per 10,000 people', caption = "Source: New York Times COVID-19 data") +
  scale_fill_continuous(labels = comma) +
  theme(legend.position="bottom") +
  scale_fill_viridis_c(option = "plasma",labels = comma)
  
```

And now we'll map out congregations per 10,000 residents:

```{r}

geo_county_covid_religion %>%
  ggplot(aes(fill = congregations_per_10k)) +
  geom_sf(lwd = 0) +
  theme_map() +
  labs(fill='Congregations per 10,000 residents',caption = "Source: The Association of Religion Data Archives, 2010 survey") +
  scale_fill_continuous(labels = comma) +
  theme(legend.position="bottom") +
  scale_fill_viridis_c(option = "plasma",labels = comma)
  
```

Still more congregations per 10K in the South/Plains areas -- less on the coasts and in major population centers.

Now we're going to plot religious adherence percentage against peak confirmed cases (per 10,000 residents).

```{r}

ggplot(county_covid_religion) +
  geom_point(aes(congregations_per_10k, cases_per_10k)) +
  labs(x="Religious congregations per 10,000 people", y="Peak COVID cases per 10,000 people", caption = "Source: NYT COVID-19 data and the Association of Religion Data Archives", fill="key") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_smooth(aes(congregations_per_10k, cases_per_10k), method = "lm", se = FALSE)

```

We can see a clearer pattern here than with the adherents -- it's looking like there are actually higher COVID instances in counties with fewer congregations. Let's 
test the relationship:

```{r}

# Test the relationship
correlation_covid_religion <- county_covid_religion %>%
  ungroup() %>%
  select(congregations_per_10k, cases_per_10k) 

cor.test(correlation_covid_religion$congregations_per_10k, correlation_covid_religion$cases_per_10k)

```

So it's not a strong correlation, but it is statistically significant. Counties where there are fewer congregations per 10,000 residents are a little bit more likely to have already experienced coronavirus outbreaks. That makes some sense; fewer congregations means that the religious adherents in a given county are more concentrated into the same worship spaces.

## Looking to the future

So maybe there's not a large correlation between religious adherence or congregations per 10,000 residents and confirmed COVID-19 cases to date. But that also makes some sense; as [The Washington Post reported] (https://www.washingtonpost.com/religion/2020/04/17/cell-phone-data-coronavirus-churchgoers-home-easter/), mobile location data shows that stay-at-home orders were actually pretty effective at keeping churchgoers home on Easter -- one of the biggest attendance days of the year for Christian churches. But some states are starting to lift their stay-at-home orders, so is it possible that some counties could be vulnerable to new outbreaks once people start to return to churches?

Let's start by using what we've already looked at to identify a subset of counties: ones with high rates of religious adherence and lower numbers of congregations per 10,000 residents. How many of the ~2,800 counties in our current dataset are in the top 20% of religious adherents and the bottom 20% of congregations per 10K?

```{r}

high_adherence_low_congregation <- county_covid_religion %>%
  filter(religious_adherence >= quantile(county_covid_religion$religious_adherence, probs = 0.80)) %>%
  filter(congregations_per_10k <= quantile(county_covid_religion$congregations_per_10k, probs = 0.20)) %>%
  select(fips, county, state, population, religious_adherence, congregations_per_10k, key) %>%
  arrange(desc(population))

high_adherence_low_congregation

```

There are some interesting spots here among the 34 results. We find several counties in Louisiana and surrounding New York City (Suffolk and Nassau on Long Island, Westchester to the immediate north, and Bergen and Union in New Jersey), two areas that are existing hot spots. We also find Fulton County in Georgia, which is home to the city of Atlanta -- and in a state that has already started reopening.

But which of these areas could be vulnerable as the hot spots of the *future*? Let's see what happens when we limit that list to places that are not already among the counties with the highest rates of COVID-19 cases:

```{r}

high_adherence_low_congregation <- county_covid_religion %>%
  filter(religious_adherence >= quantile(county_covid_religion$religious_adherence, probs = 0.80)) %>%
  filter(congregations_per_10k <= quantile(county_covid_religion$congregations_per_10k, probs = 0.20)) %>%
  filter(cases_per_10k <= quantile(county_covid_religion$cases_per_10k, probs = 0.85)) %>%
  select(county, state, population, religious_adherence, congregations, congregations_per_10k, fips, key) %>%
  arrange(desc(religious_adherence))

high_adherence_low_congregation

```

**FINDINGS BASED ON THE COUNTIES THAT RESULT FROM THIS ANALYSIS**
